\appendix

\chapter{Installation Guidelines}
All the files for this thesis can be found at \url{https://github.com/stevna/MScThesis_coopMTD}. This includes the modified version of Bashlite, all files related to the MTD framework and techniques, the scripts for data collection, and the Jupyter notebooks for data analysis. There is also a link to download all three complete virtual machines.

The Wiki tab of this GitHub page contains detailed installation and execution instructions, including how to run Bashlite and what scripts need to be executed to start the infection and mitigation processes. It also includes instructions on how to insert the downloaded virtual machines into Oracle VM VirtualBox so that the virtual machines do not need to be recreated for future research. 

\chapter{Contents of the GitHub Repository}
The provided GitHub repository has the following contents:
\begin{enumerate}
    \item This thesis as a PDF.
    \item A short summary of the thesis in German. 
    \item The modified server and client code of Bashlite.
    \item All files related to the MTD framework and techniques of all 3 virtual machines.
    \item The files and scripts to collect the data for the evaluation.
    \item The files and scripts to analyze the data for the evaluation.
    \item The Latex source code used for this report,
    \item The source files of the graphics and tables used for this thesis which is a Microsoft PowerPoint document.
    \item A SWITCHdrive link where all three virtual machines can be downloaded for the subsequent insertion into Oracle VM VirtualBox. 
    
\end{enumerate}

